### 2017-04-28 paper/jfp-submission/evaluation.tex
misc
misc tweaks

\begin{figure}[htp]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.5\linewidth]{distrib.png}
\end{minipage}
}
\vspace{3ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 76--82\% of
  the programs in under one second, improving to 84--85\% in under
  10 seconds. In both datasets we detect actual type errors at least 75\%
  of the time, unbound variables or constructors 4--5\% of the time, and
  diverging loops 3\% of the time. For the remaining 13--16\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for over 75\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we reach 84--85\% coverage, and if we allow a 60 second search,
we hit a maximum of 84--87\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 4--5\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 3\% triggered an infinite
recursion error.
%
For the remaining 13--16\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after one minute.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely uncover more errors, our experiments suggest that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{trace_size_step.png}
\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
\end{minipage}
}
\vspace{3ex}
\caption{Complexity of the generated traces. Over 80\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure}
%
\paragraph{Results}
\label{sec:results-complexity}
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions per trace is 30 for the
\ucsdbench\ dataset (45 for the \uwbench\ dataset) with a maximum of
2,745 (\resp 982) and a median of 14.5 (\resp 14.5).
%
The average number of jumps per trace is 7 (\resp 10) with a
maximium of 353 (\resp 221) and a median of 4 (\resp 4).
%
In both datasets 80\% or more traces have at most 10 jumps.

@@

\paragraph{Example: Recursion with Bad Operator}
The recursive function @sqsum@ should square each
element of the input list and then compute the sum
of the result.
%
\begin{ecode}
  let rec sqsum xs = match xs with
    | [] -> 0
    | h::t -> (*@\hlOcaml{\hlSherrloc{sqsum t}}@*) @ (h * h)
\end{ecode}
%
Unfortunately the student has used the list-append
operator |@| instead of \texttt{+} to compute the sum.
%
Both \ocaml\ and \sherrloc\ blame the \emph{wrong location},
the recursive call @sqsum t@, with the message
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
\toolname\ produces a trace showing how the evaluation of
@sqsum [1]@ gets stuck.
%
\begin{center}
  \includegraphics[height=125px]{sqsum.png}
\end{center}
%
The trace highlights the entire stuck term
(not just the recursive call), emphasizing
the \emph{conflict} between @int@ and @list@
rather than assuming one or the other is correct.

@@

\begin{figure}[t]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{user-study-reason.png}
\includegraphics[width=0.5\linewidth]{user-study-fix.png}
\end{minipage}
}
\vspace{3ex}
\caption{A classification of students' explanations and fixes for type
  errors, given either \ocaml's error % message
  or \toolname's
  jump-compressed trace. The students given \toolname's jump-compressed
  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
  error.}
\label{fig:results-user-study}
\end{figure}

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors:
%
(1) with high coverage in a timespan amenable to compile-time analysis;
%
(2) with traces that have a low average complexity of 7 jumps;
%
(3) that are more helpful to novice programmers than traditional type
error messages; and
%
(4) that can be used to automatically locate the source of a type error.

There are, of course, drawbacks to our approach. Four that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) dealing with an explosion in the size of generated traces;
%
%(3) the inability to handle certain instances of infinite types; and
(3) our use of a non-parametric function type; and
%
(4) handling ad-hoc polymorphism.

@@

\paragraph{Non-Parametric Function Type}
As we discussed in \S~\ref{sec:how-safe} some ill-typed programs
lack a witness in our semantics due to our use of a non-parametric
type $\tfun$ for functions.
%
These programs cannot ``go wrong'', strictly speaking, but would be very
difficult to \emph{use} in practice.
%
We also note that many of these programs induce cyclic typing constraints,
causing infinite-type errors, which in our experience can be particularly
difficult to debug (and to explain to novices).

Better support for these programs would be welcome.
%
For example, we might track how the types of inputs change between
recursive calls.
%
If we cannot find a traditional witness, we could then produce a trace
expanded to show these particular steps.

>>>>

\begin{figure}[t]
% \centerline{
% \begin{minipage}{1.2\textwidth}
\centering
\includegraphics[width=0.7\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
% \includegraphics[width=0.6\linewidth]{distrib.png}
% \end{minipage}
% }
% \vspace{3ex}
\caption{Results of our coverage testing. Our random search successfully
  finds witnesses for 76--82\% of the programs in under one second,
  improving to 84--85\% in under 10 seconds. }
\label{fig:results-witness}
\end{figure}
\begin{figure}[t]
\includegraphics[width=0.7\linewidth]{distrib.png}
\caption{Distribution of test outcomes. In both datasets we detect
  actual type errors at least 75\% of the time, unbound variables or
  constructors 4--5\% of the time, and diverging loops 3\% of the
  time. For the remaining 13--16\% of the programs we are unable to
  provide any useful feedback. }
\label{fig:results-distrib}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figures~\ref{fig:results-witness}~and~\ref{fig:results-distrib}.
%
In both datasets our tool was able to find a witness for over 75\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we reach 84--85\% coverage, and if we allow a 60 second search,
we hit a maximum of 84--87\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 4--5\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 3\% triggered an infinite
recursion error.
%
For the remaining 13--16\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after one minute.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely uncover more errors, our experiments suggest that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

% \item \ES{others?}
%
\begin{figure}[t]
% \centerline{
% \begin{minipage}{1.2\textwidth}
\includegraphics[width=0.7\linewidth]{trace_size_step.png}
\includegraphics[width=0.7\linewidth]{trace_size_jump.png}
% \end{minipage}
% }
% \vspace{3ex}
\caption{Complexity of the generated traces. Over 80\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure}
%
\paragraph{Results}
\label{sec:results-complexity}
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions per trace is 30 for the
\ucsdbench\ dataset (45 for the \uwbench\ dataset) with a maximum of
2,745 (\resp 982) and a median of 14.5 (\resp 14.5).
%
The average number of jumps per trace is 7 (\resp 10) with a
maximium of 353 (\resp 221) and a median of 4 (\resp 4).
%
In both datasets 60\% of traces have at most 5 jumps, and 80\% or more
have at most 10 jumps.

@@

\paragraph{Example: Recursion with Bad Operator}
The recursive function @sqsum@ should square each
element of the input list and then compute the sum
of the result.
%
\begin{ecode}
  let rec sqsum xs = match xs with
    | [] -> 0
    | h::t -> (*@\hlOcaml{\hlSherrloc{sqsum t}}@*) @ (h * h)
\end{ecode}
%
Unfortunately the student has used the list-append
operator |@| instead of \texttt{+}. % to compute the sum.
%
Both \ocaml\ and \sherrloc\ blame the \emph{wrong location},
the recursive call @sqsum t@, with the message
%
\begin{verbatim}
  This expression has type
    int
  but an expression was expected of type
    'a list
\end{verbatim}
%
\begin{center}
  \includegraphics[height=125px]{sqsum.png}
\end{center}
%
\toolname\ produces the above trace showing how the evaluation of
@sqsum [1]@ gets stuck.
%
The trace highlights the entire stuck term
(not just the recursive call), emphasizing
the \emph{conflict} between @int@ and @list@
rather than assuming one or the other is correct.

@@

\begin{figure}[t]
% \centerline{
% \begin{minipage}{1.2\textwidth}
\includegraphics[width=0.7\linewidth]{user-study-reason.png}
\includegraphics[width=0.7\linewidth]{user-study-fix.png}
% \end{minipage}
% }
% \vspace{3ex}
\caption{A classification of students' explanations and fixes for type
  errors, given either \ocaml's error % message
  or \toolname's
  jump-compressed trace. The students given \toolname's jump-compressed
  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
  error.}
\label{fig:results-user-study}
\end{figure}

@@

To summarize, our experiments demonstrate that \nanomaly finds witnesses
to type errors:
%
(1) with high coverage in a timespan amenable to compile-time analysis;
%
(2) with traces that have a low median complexity of 5 jumps;
%
(3) that are more helpful to novice programmers than traditional type
error messages; and
%
(4) that can be used to automatically locate the source of a type error.

There are, of course, drawbacks to our approach. Four that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) dealing with explosions in the size of generated traces;
%
%(3) the inability to handle certain instances of infinite types; and
(3) our use of a non-parametric function type; and
%
(4) handling ad-hoc polymorphism.

@@

\paragraph{Non-Parametric Function Type}
As we discussed in \S~\ref{sec:how-safe} some ill-typed programs
lack a witness in our semantics due to our use of a non-parametric
type $\tfun$ for functions.
%
These programs cannot ``go wrong'', strictly speaking, but would be very
difficult to \emph{use} in practice.
%
We also note that many of these programs induce cyclic typing constraints,
causing infinite-type errors, which in our experience can be particularly
difficult to debug (and to explain to novices).
%
Better support for these programs would be welcome.
%
For example, we might track how the types of inputs change between
recursive calls.
%
If we cannot find a traditional witness, we could then produce a trace
expanded to show these particular steps.
